Spfy: distributed predictive genomics of E.coli with graph based result linkage

### 2017-08-11 paper-webserver.tex
ADD: matt's comments REMOVE: a couple of section headings
ADD: talking pts for Functionality section

% surveillance
Until recently, Public-health agencies relied on laboratory tests such as XYZ to characterize bacterial isolates in outbreak and surveillance settings.
The previous gold-standard in determining strain relatedness was pulsed-field gel electrophoresis (PFGE) {ronholm2016navigating}, which uses rare-cutting restriction enzymes to produce a unique banding pattern for each strain.
However, in \textit{Enterococcus faecium}, PFGE has been shown to misclassify 9 of 132 isolates, when compared to whole-genome sequencing (WGS) based discrimination \cite{pinholt2015multiple}.
In \textit{Klebsiella pneumoniae} \cite{marsh2015genomic}, \textit{Yersinia enterocolitica} \cite{gilpin2014limitations}, and \textit{Staphylococcus aureus} \cite{doi:10.1093/ofid/ofu096}, WGS was used to discriminate isolates after initial clustering by PFGE resulted in indistinguishable samples.
Examination of PFGE bands are also subjective, difficult to share \cite{lytsy2017time}, and collative platforms such as PulseNet reported \cite{gilpin2014limitations} that even after collecting 72\% of \textit{Campylobacter jejuni} in a given year in Minnesota (673 cases), 87\% of isolates could not be linked by PFGE pattern.
To further characterize strains ... serotyping, which detects the presence of specific cell surface antigens.
Antimicrobial resistance testing, virulence factor testing ...
However, current efforts are focused on predictive genomics, where the relevant phenotypic information can be determined through examination of the whole-genome sequence.

@@

% Solution via software engineering
Current programs that address these issues include the Comprehensive Antibiotic Resistance Database (CARD) \cite{mcarthur2013comprehensive}, which addresses data concerns by curating a set of known AMR genes.
In addition, GenomeTrackr from the Food and Drug Administration (FDA) \cite{allard2016practical}; GenBank from the National Center for Biotechnology Information (NCBI) \cite{doi:10.1093/nar/gks1195}; and PulseNet from the Center for Disease Control and Prevention (CDC) \cite{swaminathan2001pulsenet}, all enable sharing of genome sequences and associated metadata, either openly or amongst member institutions.

@@

% rewrite of bigdata approach (BIGGER bigdata)
Bacterial species, pan-genome ...Panseq \cite{laing2010pan}, Roary \cite{page2015roary}, chewBBACA \url{https://github.com/mickaelsilva/chewBBACA}, and Graphtyper \cite{Eggertsson148403} can compute these pan-genomic regions, and are freely available.
These software programs are capable of processing X number of genomes per Y days at a cost of Z, which would take a A, B and C to do in a wet-lab.

@@

\section{METHODS AND IMPLEMENTATION}
% para
% goals: predictive analylitcs in minutes, storage up 50,000-100,000 genomes

@@

\subsection{Graph databases and ontologies}
% para
% 0. Goals: big-data, everything linked, easy addition of new links
% 1. spfy is built around graph technologies
% 2. graph dbs store data as a series of links
% 3. existing graph dbs and bioinformatics implementations
% para
% 1. how we structure our graph
% 2. ontoogies used
% 3. inferencing
% 3 1/2. SPARQL queries
% 4. results/timings: scale

@@

\subsection{Web design}
% para
% 1. goals: intuitive/familar, ease of use 
% 2. design specs and different examples
% 3. Google Material design
% 4. some focus on card based design
% para
% 1. implementation: reactjs, react-md, ES6, JSX
% 2. A to B: example of how users go from uploading files to results
% 3. backwards compatibility with older browsers: transpile via Babel
% 4. separation from Flask layer

@@

\subsection{Real-time analysis pipelines and task distribution}
\subsection{Real-time analysis pipelines}
% para
% 1. goals: real-time, support for pipelines (linked modules)
% 2. how pipelines have been handled in the past: Galaxy, other examples
% 3. Python, RQ
% para
% 1. how we implemented RQ
%	related: packaging of modules in conda
% 2. different pipelines available
% 3. how pipelines are started from the front-end
% 4. result storage in graph db

@@

\subsubsection{Task distribution and parallelization}
% para
% 1. goals: scale analyses to "big-data", error handling
% 2. examples of different ways parallelization has been handled in the past
% 3. how we handle parallelization with RQ
% para
% 1. how many tasks have we tested this with
% 2. error handling: rq-dashboard, sentry
% 3. why options like sentry are better than traditional logging: scales well to tons (big-data levels) of tasks, groups the same errors together, reporting via email

\subsection{Containerization}
% para
% 1. goals: why compartmentilizations
% 2. examples of different virtualization technologies
% 3. intro do docker
% para
% 1. how we implemented docker
% 2. use of docker-compose
% 3. how this lets us replicate worker containers and link everything together

@@

\subsection{Continuous integration and testing}
% para
% 1. goals: why CI, testing is important
% 2. examples of different testing platforms
% 3. how we've implemented it, integration with github

@@

\subsection{Guides and documentation}
% para
% 1. goals: use guide (ie. the README), dev guide
% 2. how docs have been done in the past (in general)
% 3. our use of ReadtheDocs

\section{RESULTS AND DISCUSSION}

\subsection{Further containerizing analyses modules}

% ex. in python, tox can be set to spin up docker containers for testing - see certbot
% could spin up a new container for each analysis module to run see \cite{di2015impact} - neglible performance impact

@@

>>>>

% surveillance
Until recently, Public-health agencies relied on laboratory tests such as XYZ to characterize bacterial isolates in outbreak and surveillance settings.
The previous gold-standard in determining strain relatedness was pulsed-field gel electrophoresis (PFGE) {ronholm2016navigating}, which uses rare-cutting restriction enzymes to produce a unique banding pattern for each strain.
However, in \textit{Enterococcus faecium}, PFGE has been shown to misclassify 9 of 132 isolates, when compared to whole-genome sequencing (WGS) based discrimination \cite{pinholt2015multiple}.
In \textit{Klebsiella pneumoniae} \cite{marsh2015genomic}, \textit{Yersinia enterocolitica} \cite{gilpin2014limitations}, and \textit{Staphylococcus aureus} \cite{doi:10.1093/ofid/ofu096}, WGS was used to discriminate isolates after initial clustering by PFGE resulted in indistinguishable samples.
Examination of PFGE bands are also subjective, difficult to share \cite{lytsy2017time}, and collative platforms such as PulseNet reported \cite{gilpin2014limitations} that even after collecting 72\% of \textit{Campylobacter jejuni} in a given year in Minnesota (673 cases), 87\% of isolates could not be linked by PFGE pattern.
Antimicrobial resistance testing, virulence factor testing ...
However, current efforts are focused on predictive genomics, where the relevant phenotypic information can be determined through examination of the whole-genome sequence.

@@

% Solution via software engineering

In addition, GenomeTrackr from the Food and Drug Administration (FDA) \cite{allard2016practical}; GenBank from the National Center for Biotechnology Information (NCBI) \cite{doi:10.1093/nar/gks1195}; and PulseNet from the Center for Disease Control and Prevention (CDC) \cite{swaminathan2001pulsenet}, all enable sharing of genome sequences and associated metadata, either openly or amongst member institutions.

@@

@@

% Some general comments
% The NAR Database issue is more of a showcase then a rigorous exploration of software design choices.
% Given this focus, i would suggest the following:
% 1. Increase/highlight the discriptions of the functions and capabilities, maybe by adding a Functionality section
% 2. In the Implmentation (or Methods) section, only give a cursory description of the layout and components. Don't need to provide too much justification
% 3. Use the Results section to highlight the scope/size and speed. This can be short
% 4. In the Discussion, this is where i would expand on the justications and reasons for specific design choices. Pick 2-3 main ones and discuss those (i.e. don't need to justify our choice of documentation software). Also compare with other software in Discussion.
% 5. Add a conclusions section

\section{FUNCTIONALITY}

% Describe available functions in spfy
% para covering everything
Spfy performs reference laboratory tasks: O-antigen and H-antigen typing, Shiga-toxin (STX) typing,  virulence factor (VF) and antimicrobial resistance (AMR) gene determination, and related strain determination.
Spfy also performs bioinformatics analyses: pangenome generation, statistical significance testing of genome markers for user-defined groups, and AMR predictions using support vector machines (SVMs).

% para covering ectyper & RGI
To further characterize strains ... serotyping, which detects the presence of specific cell surface antigens.
O-antigen and H-antigen typing, and VF determination use our ECTyper \url{https://github.com/phac-nml/ecoli_serotyping} software.
ECTYper compares the user-submitted genomes against a curated set of sequences, an approach originally implemented in the VirulenceFinder \cite{joensen2014real} tool.
The Comprehensive Antibiotic Resistance Database (CARD) \cite{mcarthur2013comprehensive} maintains a set of known AMR genes and develops the Resistance Gene Identifier (RGI) program.
Both ECTyper and RGI are based off of BLAST \cite{pmid2231712}.

% para on phylotyper
Stx typing looks at ...
Phylotyper constructs phylogenetic trees and uses ancestral reconstruction to determine Stx-types ... others.
The software can also be used to generate trees and determine closely related strains.

% para on pangenome
Bacterial species, pan-genome ...Panseq \cite{laing2010pan}, Roary \cite{page2015roary}, chewBBACA \url{https://github.com/mickaelsilva/chewBBACA}, and Graphtyper \cite{Eggertsson148403} can compute these pan-genomic regions, and are freely available.
Spfy integrates panseq ...

% storage of results
The results from ECTyper, RGI, Phylotyper, and Panseq are converted into a graph structure and linked the the genomes they originated from.

% use of stored results
By retrieving results from the database, Spfy provides group comparisons using Fishers, and can build SVMs to predict AMR based off of the presence of particular pangenome regions.

\section{IMPLEMENTATION}
% para
% goals: predictive analylitcs in minutes, storage up 50,000-100,000 genomes {}

@@

\subsection{Data Storage}
% para
% 0. Goals: big-data, everything linked, easy addition of new links
% 1. spfy is built around graph technologies
% 1. how we structure our graph
% 2. ontoogies used
% 3. inferencing
% 3 1/2. SPARQL queries

@@

\subsection{Web design}
% para
% 1. goals: intuitive/familar, ease of use 
% 2. design specs
% 3. Google Material design
% 1. implementation: reactjs, react-md, ES6, JSX
% 4. separation from Flask layer

@@

\subsection{Real-time analysis pipelines}
% para
% 1. goals: real-time, support for pipelines (linked modules)
% 2. how pipelines have been handled in the past: Galaxy, other examples
% 3. Python, RQ
% para
% 1. how we implemented RQ
%	related: packaging of modules in conda

@@

% para
% 1. goals: scale analyses to "big-data", error handling
% 3. how we handle parallelization with RQ
% para
% 1. how many tasks have we tested this with
% 2. error handling: rq-dashboard, sentry
% 3. why options like sentry are better than traditional logging: scales well to tons (big-data levels) of tasks, groups the same errors together, reporting via email

% para
% 1. goals: why compartmentilizations
% 1. how we implemented docker
% 3. how this lets us replicate worker containers and link everything together

@@

\subsection{Continuous integration and testing}
% Keep this short
% para
% 1. goals: why CI, testing is important
% 2. how we've implemented it, integration with github

@@

\section{RESULTS}

% e.g.
% database statistics
% analysis run-time / throughput with different levels of parallelization

\section{DISCUSSION}

% Mention where users / developers can find documentation

@@

\section{CONCLUSIONS}
